---
title: "DGQGSI Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DGQGSI Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

`DGQGSI` provides tools for **multi-trait selection in plant breeding programs**
using desired-gain–based selection indices. It integrates two complementary
workflows:

- **DGSI** — Desired-Gain Selection Index Optimization: estimates index weights
  whose realized genetic gains closely match breeder-defined targets
  (Joukhadar et al., 2024).
- **QGSI** — Desired-Gain Quadratic Genomic Selection Index: a nonlinear index
  that scores genotypes using both linear and quadratic GEBV components,
  capturing trait interactions that a linear index would miss
  (Cerón-Rojas et al., 2026).

The key advantage of both methods is that breeders specify **desired genetic
gains** directly — such as *"increase yield by 1.5 units while reducing disease
score by 1 unit"* — rather than economic weights, which are rarely known with
precision in practice.

---

## Methodological background

### DGSI

The DGSI finds a weight vector **b** such that the linear index

$$I = \mathbf{b}^\top \mathbf{X}$$

produces realized genetic gains **g** as close as possible to the breeder's
desired gain vector **d**. The algorithm in
`run_desired_gain_index_Joukhadar2024()`:

1. Samples candidate desired-gain vectors in the neighbourhood of **d**.
2. Computes index coefficients **b** for each candidate.
3. Evaluates the realized gains **g** on the selected subset.
4. Retains the coefficients that minimise the deviation between **g** and **d**.

Because the search is stochastic, multiple independent replicates (`n_rep`) are
run and the best solution across replicates is returned.

### QGSI

For genotype $i$ with GEBV vector $\boldsymbol{\gamma}_i$, the quadratic index
is:

$$\text{QGSI\_DG}_i = \mathbf{d}^\top \boldsymbol{\gamma}_i +
\boldsymbol{\gamma}_i^\top \mathbf{W}_d \, \boldsymbol{\gamma}_i$$

where **d** is the desired gain vector and $\mathbf{W}_d$ is the quadratic
interaction matrix built automatically by `construct_Wd_from_dg()`. The
**linear term** rewards GEBVs aligned with the desired direction; the
**quadratic term** additionally rewards genotypes that combine favourable
trait values simultaneously, capturing synergies and penalising trade-offs
invisible to a linear index.

---

## Example data

The package ships with two built-in datasets. `example_pheno.csv` contains
observed phenotypic means per genotype; `example_gebv.csv` contains genomic
estimated breeding values (GEBVs) from a genomic prediction model.

We work with six traits across two contrasting groups:

| Trait | Description                      | Direction         |
|-------|----------------------------------|-------------------|
| YLD   | Grain yield                      | higher is better  |
| MY    | Market yield                     | higher is better  |
| MI    | Marketable index                 | higher is better  |
| BL    | Bacterial leaf blight score      | lower is better   |
| NBL   | Neck bacterial leaf blight score | lower is better   |
| VHB   | Vascular hypersensitive blight   | lower is better   |

The **desired gain vector** `dg` encodes the target improvement per selection
cycle. For the three disease traits, the value represents the intended
*reduction*; direction flipping is managed internally by the `lower_is_better`
argument, so all values in `dg` are given as positive quantities regardless of
the trait direction.

```{r}
library(DGQGSI)
library(data.table)

ext   <- system.file("extdata", package = "DGQGSI")
pheno <- fread(file.path(ext, "example_pheno.csv"))
gebv  <- fread(file.path(ext, "example_gebv.csv"))

traits <- c("YLD", "MY", "MI", "BL", "NBL", "VHB")

dg <- c(YLD = 1.5, MY = 0.5, MI = 0.5, BL = 1, NBL = 1, VHB = 1)
```

---

## Run the full pipeline

`run_dgsi_qgsi_pipeline()` is the main entry point. Setting `mode = "both"`
runs DGSI and QGSI in sequence and, with `merge_outputs = TRUE`, combines their
genotype rankings into a single comparison table via `compare_dg_and_qgsi()`.

Key arguments:

| Argument             | Description                                                      |
|----------------------|------------------------------------------------------------------|
| `mode`               | `"dg"`, `"qgsi"`, or `"both"`                                   |
| `init_data`          | Genotype IDs and metadata to preserve in output                  |
| `cand_data`          | Candidate phenotypic trait values (DGSI input)                   |
| `ref_data`           | Reference population for scaling and covariance estimation       |
| `gebv_data`          | Genotype-level GEBVs (QGSI input)                                |
| `trait_cols`         | Character vector of trait column names                           |
| `dg`                 | Named numeric vector of desired gains                            |
| `lower_is_better`    | Traits for which a smaller value is the breeding goal            |
| `trait_min_sd`       | Per-trait minimum threshold for threshold-based selection        |
| `n_iter`             | Optimisation iterations per DGSI replicate                       |
| `n_rep`              | Number of independent DGSI replicates                            |
| `dg_scale_traits`    | Whether to center and scale traits before DGSI                   |
| `qgsi_center_traits` | Whether to center GEBVs before QGSI                              |
| `qgsi_scale_traits`  | Whether to scale GEBVs before QGSI                               |
| `merge_outputs`      | Whether to merge DGSI and QGSI rankings into one table           |

```{r}
res <- run_dgsi_qgsi_pipeline(
  mode = "both",
  init_data  = pheno[, .(GenoID, Family)],
  cand_data  = pheno[, c("GenoID", traits), with = FALSE],
  ref_data   = pheno[, c("GenoID", traits), with = FALSE],
  gebv_data  = gebv,
  trait_cols = traits,
  dg         = dg,
  lower_is_better = c("BL", "NBL", "VHB"),
  trait_min_sd    = c(YLD = 0.2, MY = 0.1, MI = 0.1,
                      BL  = 0.1, NBL = 0.1, VHB = 0.1),
  dg_scale_traits    = FALSE,
  qgsi_center_traits = FALSE,
  qgsi_scale_traits  = FALSE,
  n_iter        = 50,
  n_rep         = 3,
  merge_outputs = TRUE,
  debug         = FALSE
)
```

The pipeline returns a named list. The components present depend on `mode`:

- `res$dg_result` — full DGSI output from
  `run_desired_gain_index_Joukhadar2024()`.
- `res$W_d_result` — the automatically constructed $\mathbf{W}_d$ matrix
  from `construct_Wd_from_dg()`.
- `res$qgsi_result` — full QGSI output from `run_qgsi_desired_gain()`.
- `res$comparison_result` — merged comparison table from
  `compare_dg_and_qgsi()` (only when `mode = "both"` and
  `merge_outputs = TRUE`).

---

## Exploring the results

### The comparison table

The merged comparison table is the primary output for side-by-side evaluation.
It contains, per genotype: the DGSI score (`DG_SelectionIndex`), the DGSI
selection flag (`DG_Selected`), the linear and quadratic QGSI components
(`LinearDGPart`, `QuadraticDGPart`), the final QGSI score (`QGSI_DG`), the
QGSI rank (`Rank_QGSI_DG`), and a sign-agreement flag (`SignSame_DG_QGSI`).

```{r}
comp <- res$comparison_result$comparison_table
head(comp)
```

### Top candidates by DGSI

Sorting by `DG_SelectionIndex` (descending) reveals the genotypes whose
multi-trait phenotypic profile best matches the optimised linear index weights:

```{r}
head(comp[order(-DG_SelectionIndex)], 10)
```

### Top candidates by QGSI

Sorting by `Rank_QGSI_DG` (ascending) highlights genotypes that combine
favourable GEBVs across multiple traits simultaneously. The quadratic term
rewards candidates that are good on several fronts at once:

```{r}
head(comp[order(Rank_QGSI_DG)], 10)
```

### Rank agreement between DGSI and QGSI

A Spearman correlation between the DGSI score ranking and the QGSI rank
summarises overall agreement across all candidates:

```{r}
cor_val <- cor(
  rank(-comp$DG_SelectionIndex),
  comp$Rank_QGSI_DG,
  method = "spearman"
)
cat(sprintf("Spearman rank correlation (DGSI vs QGSI): %.3f\n", cor_val))
```

A high correlation indicates both methods largely agree on which genotypes to
advance. When the correlation is lower, genotypes that score high on one method
but not the other deserve closer inspection: they tend to have strong trait
interactions that only the quadratic term captures.

The Pearson and Spearman correlations computed by `compare_dg_and_qgsi()` are
also available directly:

```{r}
res$comparison_result$correlation_summary
```

---

## Inspecting individual workflow outputs

### DGSI output

The DGSI result contains several diagnostics beyond the ranked genotype table:

```{r}
# Optimised index coefficients
res$dg_result$optimized_b

# Realized gains in the transformed trait space
res$dg_result$realized_g

# Objective value of the best replicate (lower = better match to dg)
res$dg_result$best_q

# Mean pairwise correlation among replicate index vectors (stability check)
res$dg_result$mean_replicate_cor

# Trait-level summary: all candidates vs. selected subset
res$dg_result$selection_summary
```

The `selection_summary` table reports the mean and SD for each trait in the
full candidate set and in the selected subset, together with the realized gain
expressed in SD units. Values above 0.9 for `mean_replicate_cor` indicate good
agreement across replicates.

### QGSI output

The QGSI result separates the linear and quadratic components per genotype:

```{r}
head(res$qgsi_result$ranked_geno)
```

A component-level summary (mean, SD, range) is also available:

```{r}
res$qgsi_result$component_summary
```

---

## Running each method separately

Both workflows can also be called independently when finer control is needed.

### DGSI only

`run_desired_gain_index_Joukhadar2024()` supports two selection strategies via
`select_mode`:

- `"top_n"` — selects the `n_select` highest-ranking genotypes.
- `"trait_thresholds"` — selects genotypes meeting minimum per-trait thresholds
  (`trait_min_sd`), with optional fallback to top-N if no genotype passes.

```{r, eval=FALSE}
dgsi_res <- run_desired_gain_index_Joukhadar2024(
  init_data       = pheno[, .(GenoID, Family)],
  cand_data       = pheno[, c("GenoID", traits), with = FALSE],
  ref_data        = pheno[, c("GenoID", traits), with = FALSE],
  trait_cols      = traits,
  dg              = dg,
  lower_is_better = c("BL", "NBL", "VHB"),
  select_mode     = "trait_thresholds",
  trait_min_sd    = c(YLD = 0.2, MY = 0.1, MI = 0.1,
                      BL  = 0.1, NBL = 0.1, VHB = 0.1),
  scale_traits    = FALSE,
  n_iter          = 200,
  n_rep           = 10,
  debug           = FALSE
)

head(dgsi_res$ranked_geno)
dgsi_res$selection_summary
```

### QGSI only

`run_qgsi_desired_gain()` requires GEBVs and a desired-gain vector. A full
$\mathbf{W}_d$ matrix can be supplied via `W_d`; otherwise a diagonal matrix
is built from `quadratic_diag_weights`, which captures only squared trait
effects without cross-trait interactions.

```{r, eval=FALSE}
qgsi_res <- run_qgsi_desired_gain(
  init_data         = pheno[, .(GenoID, Family)],
  gebv_data         = gebv[, c("GenoID", traits), with = FALSE],
  trait_cols        = traits,
  dg                = dg,
  lower_is_better   = c("BL", "NBL", "VHB"),
  center_traits     = FALSE,
  scale_traits      = FALSE,
  return_components = TRUE,
  debug             = FALSE
)

head(qgsi_res$ranked_geno)
qgsi_res$component_summary
```

---

## The quadratic interaction matrix W_d

`construct_Wd_from_dg()` derives $\mathbf{W}_d$ from the desired gain vector
and the empirical trait correlation structure in the GEBV data. It supports
four construction strategies via the `method` argument:

| Method            | Description                                                           |
|-------------------|-----------------------------------------------------------------------|
| `"hybrid"`        | Combines diagonal, outer-product, and correlation-weighted components |
| `"diag_only"`     | Diagonal matrix only; no cross-trait interactions                     |
| `"dg_outer"`      | Outer product $\mathbf{d}\mathbf{d}^\top$ only                        |
| `"corr_weighted"` | Outer product weighted by the empirical trait correlation matrix      |

```{r}
Wobj <- construct_Wd_from_dg(
  gebv_data      = gebv[, c("GenoID", traits), with = FALSE],
  trait_cols     = traits,
  dg             = dg,
  lower_is_better = c("BL", "NBL", "VHB"),
  center_traits  = FALSE,
  scale_traits   = FALSE,
  method         = "hybrid",
  lambda_diag    = 1.0,
  lambda_outer   = 0.5,
  lambda_corr    = 1.0,
  offdiag_shrink = 0.5,
  normalize      = "trace",
  debug          = FALSE
)

round(Wobj$W_d, 4)
```

Positive off-diagonal entries indicate that simultaneous improvement in two
traits is rewarded by the quadratic term. The diagonal entries encode the
per-trait squared contribution, scaled to the magnitude of each desired gain.
The empirical trait correlation matrix used to build $\mathbf{W}_d$ is also
returned:

```{r}
round(Wobj$cor_matrix, 3)
```

When `mode = "qgsi"` or `mode = "both"` in the pipeline, $\mathbf{W}_d$ is
constructed automatically with the same default settings unless a pre-built
matrix is passed via `W_d`.

---

## Merging independent outputs

If you run DGSI and QGSI in separate calls, `compare_dg_and_qgsi()` merges
them by genotype ID and computes rank-difference diagnostics and correlation
summaries:

```{r, eval=FALSE}
merged <- compare_dg_and_qgsi(
  dg_result                = dgsi_res,
  qgsi_result              = qgsi_res,
  id_col                   = "GenoID",
  compute_rank_differences = TRUE,
  add_correlation_summary  = TRUE,
  sort_by                  = "DG",
  debug                    = FALSE
)

head(merged$comparison_table)
merged$correlation_summary
```

The `sort_by` argument accepts `"DG_rank"`, `"QGSI_rank"`, `"DG"`, `"QGSI"`,
or `"none"`. Note that `"DG_rank"` requires `Rank_SelectionIndex` to be
present in `dg_result$ranked_geno`; use `"DG"` to sort by score when running
`select_mode = "trait_thresholds"`.

---

## Practical guidance

**Trait scaling.** If traits are on very different raw scales, set
`dg_scale_traits = TRUE` for DGSI and `qgsi_scale_traits = TRUE` for QGSI.
After scaling, `dg` values are interpreted in standard-deviation units.
When traits are already comparable, leave both flags as `FALSE` (the default).

**Calibrating desired gains.** The relative magnitudes of values in `dg`
determine the emphasis placed on each trait. Calibrate targets against
historical rates of genetic gain in your germplasm to ensure they are
biologically achievable in a single selection cycle.

**`lower_is_better` traits.** Include every trait for which a decrease is the
breeding goal. Internally, the function multiplies these columns by $-1$ so
that all objectives point in the same direction. Always specify `dg` values as
positive quantities regardless of trait direction.

**Optimisation stability.** The DGSI search is stochastic. Increase `n_rep`
and `n_iter` for more stable results at the cost of computation time.
For exploratory work, `n_iter = 50` and `n_rep = 3` are adequate. For final
selection decisions, `n_iter = 200` and `n_rep = 10` or higher are recommended.
The `mean_replicate_cor` diagnostic provides a quick stability check: values
above 0.9 indicate good agreement across replicates.

**Choosing between DGSI and QGSI.**

| Situation                               | Recommended method                         |
|-----------------------------------------|--------------------------------------------|
| Initial candidate screening             | DGSI — simpler, interpretable coefficients |
| Traits are known to interact            | QGSI — captures nonlinear synergies        |
| Rank stability check desired            | Run both; inspect Spearman correlation     |
| Genomic selection programme with GEBVs  | QGSI — designed for GEBV input             |
| Phenotype-only data, no GEBVs available | DGSI                                       |

**Debug mode.** Set `debug = TRUE` in any function to print intermediate
matrices, gain vectors, and sorting decisions at each step.

---

## Citation

Akohoue F (2026). *DGQGSI: Desired-Gain Selection Index and Quadratic Genomic
Selection Index tools for breeding programs*. R package version 0.1.0.
<https://github.com/FAkohoue/DGQGSI>

---

## References

Joukhadar R, Li Y, Thistlethwaite R, Forrest KL, Tibbits JF, Trethowan R,
Hayden MJ (2024). Optimising desired gain indices to maximise selection
response. *Frontiers in Plant Science* **15**:1337388.
<https://doi.org/10.3389/fpls.2024.1337388>

Cerón-Rojas JJ, Montesinos-López OA, Montesinos-López A, et al. (2026).
Nonlinear genomic selection index accelerates multi-trait crop improvement.
*Nature Communications* **17**:1991.
<https://doi.org/10.1038/s41467-026-69890-3>
